00-OD-preparation

### 2022-04-07 00-OD-preparation.Rmd
edits to OD-preparation file and finished set-up for r5r car travel time file.

```{r}
OD_v <- read_excel("EDM2018VIAJES.xlsx") #here we import the trips file, by default this function imports the first sheet in an excel file. This .Rmd file is saved in the same place as the .xlsx file so we just need to write the name of the file we want to read. run ' help(read_excel) ` to see the documentation of this function.

@@

Or view it in RStudio:
```{r}
view(OD_v)
```

I'm interested in making the table spatial! It currently does not have coordinates but we can add from the shapefile which can be downloaded from the [Open Data Portal Madrid Regional Transport Consortium](https://datos.crtm.es/datasets/zonificacionzt1259/explore?location=40.420959%2C-3.695073%2C12.93):
```{r}
# here we download the madrid zona shapefile using the API GeoJSON resource! neat
zn <- geojson_read("https://opendata.arcgis.com/datasets/97f83bab03664d4e9853acf0e431d893_0.geojson",  what = "sp")
zn<- st_as_sf(zn) %>% st_transform(crs = 4326) #transform it into a 'sf' class object and transform coordinates to WGS84

#let's save it to data-raw just in case this link/resource doesn't work in the future
st_write(zn, "zn.shp")
```

@@

We also know that the majority do make this trip by personal car "¿Dispone de vehículo particular para hacer este viaje?" (1):
```{r}
hist(OD_v_Oc_Dt$VVEHICULO)
```
Try exploring the rest of the variables!

@@

First, we count the number of trips coming from each origin:
```{r}
Dtrips_zona <- OD_v_Oc_Dt %>% group_by(VDESZT1259)  %>% summarise(Dtrips = (n())) # we group by the origin ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.

@@

Now, we add this new column to the zonas sf object:
```{r}
zn <- zn %>% merge(Dtrips_zona, by.x = "ZT1259", by.y="VDESZT1259")

head(zn %>% st_drop_geometry(), 12) # notice the new 'Otrips' column
```

a simple plot:
```{r}
ggplot() +
  geom_sf(data = zn,
          aes(fill = Dtrips),
          color  = NA) +
  scale_fill_distiller(palette = "Spectral")
```

>>>>

```{r}
OD_v <- read_excel("data-raw/EDM2018VIAJES.xlsx") #here we import the trips file, by default this function imports the first sheet in an excel file. This .Rmd file is saved in the same place as the .xlsx file so we just need to write the name of the file we want to read. run ' help(read_excel) ` to see the documentation of this function.

@@

Or view it in RStudio:
```{r}
#view(OD_v)
```

I'm interested in making the table spatial! It currently does not have coordinates but we can add from the shapefile which can be downloaded from the [Open Data Portal Madrid Regional Transport Consortium](https://datos.crtm.es/datasets/zonificacionzt1259/explore?location=40.420959%2C-3.695073%2C12.93):
```{r}
# # here we download the madrid zona shapefile using the API GeoJSON resource! neat
# zn <- geojson_read("https://opendata.arcgis.com/datasets/97f83bab03664d4e9853acf0e431d893_0.geojson",  what = "sp")
# zn<- st_as_sf(zn) %>% st_transform(crs = 4326) #transform it into a 'sf' class object and transform coordinates to WGS84
# 
# #let's save it to data-raw just in case this link/resource doesn't work in the future
# st_write(zn, "data-raw/zn.shp")

zn <- st_read("data-raw/zn.shp")
```

@@

We also know that the majority do make this trip by personal car "¿Dispone de vehículo particular para hacer este viaje?" (1):
```{r}
hist(OD_v_Oc_Dt$VVEHICULO)

OD_v_Oc_Dt$VVEHICULO %>% typeof()
```

Let's see how long trips are (by distance):
```{r}
t <- OD_v_Oc_Dt %>% filter(DISTANCIA_VIAJE <= 60) %>% select(DISTANCIA_VIAJE) %>% unlist() %>% as.numeric() 
hist(t)
```

@@

First, we count the number of trips coming to each destination:
```{r}
Dtrips_zona <- OD_v_Oc_Dt %>% group_by(VDESZT1259)  %>% summarise(Dtrips = (n())) # we group by the Destination ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Dtrips' column.

@@

Now, we add this new column to the zonas sf object:
```{r}
zn_t <- zn %>% merge(Dtrips_zona, by.x = "ZT1259", by.y="VDESZT1259")

head(zn_t %>% st_drop_geometry(), 12) # notice the new 'Otrips' column
```

a simple plot of the number of DESTINATION home-based trips:
```{r}
ggplot() +
  geom_sf(data = zn_t,
          aes(fill = Dtrips),
          color  = NA) +
  scale_fill_distiller(palette = "Spectral")
```
For comparision... lets plot the number of ORIGIN home-based trips:
```{r}
Otrips_zona <- OD_v_Oc_Dt %>% group_by(VORIZT1259)  %>% summarise(Otrips = (n())) # we group by the origin ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.

zn_t <- zn_t %>% merge(Otrips_zona, by.x = "ZT1259", by.y="VORIZT1259", all.x = TRUE)

ggplot() +
  geom_sf(data = zn_t,
          aes(fill = Otrips),
          color  = NA) +
  scale_fill_distiller(palette = "Spectral")
```

Let's compare these two patterns if we only include trips in which the destination is within the LEZ.

<!-- to do once Javier adds the LEZ file to data-raw or replicate how the 'zn' object was imported-->

## Saving data
```{r}
save(zn, file = "data/zn.RData")
```
